\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-02-21 document/XSGRB_sample.tex
Last commit before Arxiv submission

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[longauth]{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass[referee]{aa}

@@

\title{The X-shooter GRB afterglow legacy sample (XS-GRB). \thanks{Based on
		observations collected at the European Southern Observatory, Paranal, Chile,
		Program IDs: 084.A-0260, 084.D-0265, 085.A-0009, 086.A-0073, 087.A-0055,
		088.A-0051, 089.A-0067, 090.A-0088, 091.A-0877, 0091.C-0934, 092.A-0124,
		092.D-0056, 093.A-0069, 094.A-0134, 095.A-0045, 095.B-0811, 096.A-0079,
		097.A-0036, and 098.A-0055.}}

@@

\keywords{Gamma-ray burst: general --- Galaxies: high-redshift --- ISM: general --- Techniques: spectroscopic}

@@

	\item No bright, nearby stars within $ 1.8 + 0.4 \times \exp[(20 - R)/2.05$]
arcsec, where $R$ is the USNO magnitude of the star)

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in Galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	stars/circles are the positions of all the 1266 \textit{Swift} bursts detected
	until \termdate. Stars indicate bursts with measured redshifts and circles
	indicate those without. Blue stars indicate the position of the 104 bursts
	fulfilling the sample criteria specified in Sect. \ref{samplecrit} that have a
	measured redshift. Red outlines are added to the 84 GRBs that enter our
	statistical sample with both X-shooter spectroscopy and a measured redshift.
	Red outlines of empty stars represent bursts which have been observed with
	X-shooter and has a measured redshift, but is not a part of our statistical
	sample. The blue dots show the positions of the 61 GRBs of our statistical
	sample that lack redshift measurements. A red outline is added around the six
	bursts in our statistical sample which were observed with X-shooter, but did
	not yield a redshift measurement. Two empty circles with red outlines indicate
	bursts outside the statistical sample that were followed up with X-shooter, but
	without a redshift measurement. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998}. Please note that we removed the background where the
	sample criterion is violated ($A_V > 0.5$ mag) and replaced it with a white
	background. The grey scale bar below indicates the value of $A_V$ on the plot.
	The dotted lines indicate intervals of 30$^\circ$ in longitude and latitude.}
\label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of eight years corresponding to the ESO observing periods P84 through
P98 under the following programme IDs: 084.A-0260, 085.A-0009, 086.A-0073,
087.A-0055, 088.A-0051, 089.A-0067, 090.A-0088, 092.A-0124, 093.A-0069,
094.A-0134, 095.A-0045, 096.A-0079, 097.A-0036, and 098.A-0055 (PI: Fynbo) and
0091.C-0934 (PI: Kaper). These proposals were initiated on Guaranteed Time. We
have included a few additional bursts, from the programmes 084.D-0265 (PI:
Benetti), 091.A-0877 (PI: Schady), 092.D-0056 (PI: Rau), 092.D-0633 (PI:
Greiner), and 095.B-0811 (PI: Levan). The total collection of spectra represents
\textit{all} GRB afterglows that have been followed up by X-shooter up to
\termdate, which marks the end of the XS-GRB legacy follow-up program.

@@

\begin{figure}[!t]
	\centerline{\includegraphics[width=8cm]{figures/resolutuions}} \caption{Blue
		data points show the FWHM (km/s) of Gaussian fits to unresolved telluric
		absorption lines in the VIS spectra, as a function of the airmass corrected
		DIMM seeing. The red data points show a subsample of measurements obtained for
		NIR spectra. The colored lines show the corresponding linear fits to the
		data points. The colored bounds contain 68 per cent of the best-fit probability
		mass. The red and blue dashed lines indicate the nominal resolution for the
		slit-widths used for the telluric standard observations. As can be seen, the
		effective resolution is in many cases superior to the nominal one.}
	\label{fig:res}
\end{figure}

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$R$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission lines, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have marked the most prominent
	lines seen in GRB afterglows from \citet{Christensen2011a}. The regions of most
	severe telluric absorption are highlighted by grey-shading the background.
	Additionally, three intervening systems are seen in this sightline. This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

All the spectra are made available as a single
\href{https://sid.erda.dk/cgi-sid/ls.py?share_id=DBuNORk1lI}{ZIP file}, through
\url{http://grbspec.iaa.es}, and additionally through the ESO archive in the
form of \href{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}{phase 3
	material}. This release includes both prompt afterglow observations as well as
late time observations of the associated hosts, and represents \textit{all}
afterglow spectra of GRBs carried out by the X-shooter spectrograph since the
commissioning of the instrument, \startdate, and until the end of the last
period of the program 098.A-0055, \termdate~and thus constitutes eight years of GRB
afterglow observations with X-shooter. An overview of all the spectra and their
observational setups is given in Table \ref{tab:sample_overview}. For each
burst, each individual observation is provided in a separate reduction, and in
cases where observations have been repeated for an increased signal-to-noise or
to follow the temporal evolution, a combined spectrum is also provided. No
attempt has been made to join the spectroscopic arms, so for each observation,
three spectra are provided in separate files.

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=9cm]{figures/timing.pdf}} \caption{Afterglow
	magnitude at the start of observation and redshift completeness as a function
	of follow-up delay for all the afterglows that have been followed up. The
	points have been colored based on the redshift of the corresponding burst. Red
	symbols indicate GRBs without a measured redshift and arrows indicate bursts
	for which the afterglow was not detected in the acquisition image. In red is
	shown the redshift completeness as a function of follow-up delay.}
\label{fig:timing}
\end{figure}

@@

\begin{figure*}
	\centerline{\includegraphics[width=18cm]{figures/completeness_BAT.pdf}}
\caption{Comparison between the burst properties of all bursts observed with
	\textit{Swift} (blue), the statistical sample that fullfill the criteria
	specified in Sect. \ref{samplecrit} (green), and the subset that has been
	observed as part of the statistical sample (red). The left plot shows the
	fluence in the 15-150 keV band as observed by BAT. The middle panel shows the 0.3 - 10
	keV flux, 11 hours after the bursts as measured by XRT. In the right most
	panel, we show $N_{\mathrm{HI,X}}$ based on the XRT spectrum
	\citep{Evans2009}.} \label{fig:swift_complete}
\end{figure*}

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=9cm]{figures/XSGRB_p_values.pdf}}
\caption{Relational graph showing the respective $p$-values. They all represent
	the degree that the different samples are drawn from the same underlying
	distribution. The arrows represent the comparison direction with each of the
	$p$-values they are listed next to. Only in the H{\sc I} column density distribution is
	there mild evidence against the null hypothesis, but the discrepancy is mainly
	driven by a relatively larger fractional contribution from low-column hosts in
	the statistical sample.} \label{fig:p_values}
\end{figure}

@@

\begin{figure*}[!ht]
	\centering \includegraphics[width=16cm]{figures/XSGRB_redshift.pdf}
\caption{Redshift distribution as a function of intrinsic BAT $\gamma$-ray
	energy, $E_{\mathrm{BAT}}$. Bursts that are a part of the statistical sample
	are marked by blue stars whereas black dots show all GRBs observed with
	X-shooter. All \swift~GRBs with measured redshifts are shown in gray. For
	comparison, we overplot with red stars the GRB \citep{Tanvir2009b,
		Salvaterra2009a}, quasar \citep{Banados2017}, and galaxies \citep{Zitrin2015,
		Oesch2016} with the highest spectroscopically confirmed redshifts, the latter
	three shown at arbitrary $E_{\mathrm{BAT}}$. The blue solid line represents the
	so-called 1 s BAT sensitivity limit described in the text. The estimated epoch
	of reionization is shown by the black dot-dashed line, with the uncertainty
	shown as the black dashed lines, from the most recent measurement by
	\citet{Planck2015}. On the top of the plot is shown the marginalization of the
	redshift. Again the blue histogram represents the bursts than enter our sample
	and the white histogram the full GRB population. } \label{fig:z}
\end{figure*}

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=9cm]{figures/betaOX.pdf}}
\caption{$\beta_{OX}$ against X-ray derived hydrogen column density. In red is
	shown the sample presented in \citet{Fynbo2009} where the lines indicate the
	kernel density estimate of the distribution. For the kernel density estimate,
	the limits have been replaced with values. We note that contrary to what is
	presented in \citet{Fynbo2009}, we exclude all bursts for which we do not have
	a redshift. Darker colors represent a higher density of points. In blue are the
	points for the bursts presented here along with the marginal histograms. Limits
	on $\beta_{OX}$ are shown by downwards facing arrows. The corresponding
	marginal distributions are shown along the edges of the plot. A Kendall's
	$\tau$ test, indicates a statistically significant, low degree of negative
	correlation ($\Gamma = -0.21$ at a $p$-value = 0.01)} \label{fig:betaOX}
\end{figure}

@@

\begin{figure}[!t]
	\centering \includegraphics[width=9cm]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in Tanvir et al. (submitted), along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}.} \label{fig:NH_dist}
\end{figure}

@@

\begin{acknowledgements}
Much of the work done here would not have been done without Neil Gehrels, who has now passed away. We owe him a large debt for his invaluable work with the \swift~satellite. We also want to remember Javier Gorosabel, who was taken from us too early. 
%
JPUF, BMJ and DX acknowledge support from the ERC-StG grant EGGS-278202.
The Dark Cosmology Centre was funded by the Danish National Research
Foundation. 
%
This work was supported by a VILLUM FONDEN Investigator grant to JH
(project number 16599). 
%
TK acknowledges support by the European Commission
under the Marie Curie Intra-European Fellowship Programme in FP7.  
%
LK and JJ acknowledges
support from NOVA and NWO-FAPESP grant for advanced instrumentation in
astronomy. 
%
KEH and PJ acknowledge support by a Project Grant (162948--051) from
The Icelandic Research Fund. 
%
AG acknowledges the financial support from the
Slovenian Research Agency (research core funding No. P1-0031 and project grant
No. J1-8136). 
%
CT acknowledges support from a Spanish National Research Grant of Excellence
under project AYA 2014-58381-P and funding associated to a Ramón y Cajál
fellowship under grant number RyC-2012-09984.
%
AdUP acknowledges support from a Ramón y Cajal fellowship, a BBVA Foundation
Grant for Researchers and Cultural Creators, and the Spanish Ministry of Economy
and Competitiveness through project AYA2014-58381-P.
%
ZC acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2014-21669.
%
DAK acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2015- 26153.
%
RSR acknowledges AdUP's BBVA Foundation Grant for Researchers and Cultural
Creators and support from ASI (Italian Space Agency) through the Contract n. 2015-046-R.0 and from European Union Horizon 2020 Programme under the AHEAD project (grant agreement n. 654215).
%
GL is supported by a research grant (19054) from VILLUM FONDEN.
%
SDV acknowledges the support of the French National Research Agency (ANR) under contract ANR-16-CE31-0003 BEaPro
%
DM acknowledges support from the Instrument Center for Danish Astrophysics (IDA).
%
This work made use of data supplied by the UK {\it Swift} Science Data Centre at
the University of Leicester. Finally, it is with pleasure that we acknowledge
expert support from the ESO staff at the Paranal and La Silla observatories in
obtaining these target of opportunity data.

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=16cm]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=16cm]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

These data have not been published elsewhere. Observations for GRB~161007A
started 323 hr after the burst trigger and were centered on the potential host
galaxy (see GCN 20014 \citep{GCN20014} and GCN 20020 \citep{GCN20020}). $4
\times 600$ s of observations were obtained revealing a faint continuum rising
abruptly above the noise at $\sim 685$ nm and extending up to 2100 nm. A very
low significance continuum is detected at shorter wavelengths, down to $\sim
600$ nm. No significant emission features could be identified. Two possibilities
can explain the observed continuum break, but neither are fully satisfactory.
The first option is the Lyman alpha cutoff at $z \sim 4.6$; at this redshift, the
usually detected strong nebular lines would be shifted out of the covered
wavelength range. The host would be however exceptionally bright, with an
absolute magnitude at $\approx 150$~nm{} $M = -23.8$~AB, that is about 2.8~mag
brighter than $M^*$ at that redshift (e.g. \citealt{Bouwens2015}). The inferred
X-ray hydrogen-equivalent column density would also be a record-high $4 \times
10^{21}$~cm$^{-2}$ (see e.g. \citealt{Campana2012}). Alternatively, the
continuum discontinuity could be the 400~nm{} break at $z = 0.71$; such
feature is however prominent only in early-type galaxies, which are unlikely
hosts of long-duration GRBs (but see \citealt{Rossi2014}). Given the lack of
optical variability, it cannot be excluded that the object within the XRT error
circle is a chance association, as the probability of a random alignment is
small but non negligible (of the order of 1\%). In conclusion, we cannot provide
a secure redshift for GRB 161007A.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[longauth]{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass[referee]{aa}

@@

\title{The X-shooter GRB afterglow legacy sample (XS-GRB)\thanks{Based on
		observations collected at the European Southern Observatory, Paranal, Chile,
		Program IDs: 084.A-0260, 084.D-0265, 085.A-0009, 086.A-0073, 087.A-0055,
		088.A-0051, 089.A-0067, 090.A-0088, 091.A-0877, 0091.C-0934, 092.A-0124,
		092.D-0056, 092.D-0633, 093.A-0069, 094.A-0134, 095.A-0045, 095.B-0811,
		096.A-0079, 097.A-0036, 098.A-0136, and 098.A-0055.}}

@@

\keywords{Gamma-ray burst: general --- Galaxies: high-redshift --- ISM: general --- Techniques: spectroscopic --- Galaxies: star formation --- catalogs}

@@

	\item No bright, nearby stars within $ 1.8 + 0.4 \times \exp[(20 - R)/2.05$]
arcsec, where $R$ is the USNO magnitude of the star.

@@

\begin{figure*}
	\centerline{\includegraphics[width=\linewidth]{figures/skymap.pdf}} \caption{Mollweide projection
	in Galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	stars/circles are the positions of all the 1266 \textit{Swift} bursts detected
	until \termdate. Stars indicate bursts with measured redshifts and circles
	indicate those without. Blue stars indicate the position of the 104 bursts
	fulfilling the sample criteria specified in Sect. \ref{samplecrit} that have a
	measured redshift. Red outlines are added to the 84 GRBs that enter our
	statistical sample with both X-shooter spectroscopy and a measured redshift.
	Red outlines of empty stars represent bursts which have been observed with
	X-shooter and has a measured redshift, but is not a part of our statistical
	sample. The blue dots show the positions of the 61 GRBs of our statistical
	sample that lack redshift measurements. A red outline is added around the six
	bursts in our statistical sample which were observed with X-shooter, but did
	not yield a redshift measurement. Two empty circles with red outlines indicate
	bursts outside the statistical sample that were followed up with X-shooter, but
	without a redshift measurement. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998}. Please note that we removed the background where the
	sample criterion is violated ($A_V > 0.5$ mag) and replaced it with a white
	background. The grey scale bar below indicates the value of $A_V$ on the plot.
	The dotted lines indicate intervals of 30$^\circ$ in longitude and latitude.}
\label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during the
duration of this follow-up campaign. The observations have been taken during a
period of eight years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 084.A-0260, 085.A-0009, 086.A-0073,
087.A-0055, 088.A-0051, 089.A-0067, 090.A-0088, 092.A-0124, 093.A-0069,
094.A-0134, 095.A-0045, 096.A-0079, 097.A-0036, and 098.A-0055 (PI: Fynbo) and
0091.C-0934 (PI: Kaper). These proposals were initiated on Guaranteed Time. We
have included a few additional bursts, from the programmes 084.D-0265 (PI:
Benetti), 091.A-0877 (PI: Schady), 092.D-0056 (PI: Rau), 092.D-0633,
098.A-0136 (PI: Greiner), and 095.B-0811 (PI: Levan). The total collection of
spectra represents \textit{all} GRB afterglows that have been followed up by
X-shooter up to \termdate, which marks the end of the XS-GRB legacy follow-up
program.

@@

\begin{figure}[!t]
	\centerline{\includegraphics[width=\columnwidth]{figures/resolutuions}} \caption{Blue
		data points show the FWHM (km/s) of Gaussian fits to unresolved telluric
		absorption lines in the VIS spectra, as a function of the airmass corrected
		DIMM seeing. The red data points show a subsample of measurements obtained for
		NIR spectra. The colored lines show the corresponding linear fits to the
		data points. The colored bounds contain 68 per cent of the best-fit probability
		mass. The red and blue dashed lines indicate the nominal resolution for the
		slit-widths used for the telluric standard observations. As can be seen, the
		effective resolution is in many cases superior to the nominal one.}
	\label{fig:res}
\end{figure}

@@

\begin{figure*}
	\centerline{\includegraphics[width=\linewidth]{figures/GRB121024A.pdf}}
\caption{Telluric corrected, normalized spectrum of GRB~121024A at $z = 2.300$
	that illustrates the typical data quality. The continuum estimate is shown in
	dashed red and the error spectrum in solid blue. The acquisition magnitude is
	$R$ = 20, meaning it is in the brighter end of the sample presented here, but
	not the brightest. The spectrum is rich in absorption and emission lines, including
	absorption from molecular $\mathrm{H_2}$. The absorption trough visible at
	$\sim$ 4000 $\AA$ is due to \lya~in the host. We have marked the most prominent
	lines seen in GRB afterglows from \citet{Christensen2011a}. The regions of most
	severe telluric absorption are highlighted by grey-shading the background.
	Additionally, three intervening systems are seen in this sightline. This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
\label{fig:spectrum}
\end{figure*}

@@

All the spectra are made available as a single
\href{https://sid.erda.dk/cgi-sid/ls.py?share_id=DBuNORk1lI}{ZIP file}, through
\url{http://grbspec.iaa.es} \citep{DeUgartePostigo2014f}, and additionally through the ESO archive in the
form of \href{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}{phase 3
	material}. This release includes both prompt afterglow observations as well as
late time observations of the associated hosts, and represents \textit{all}
afterglow spectra of GRBs carried out by the X-shooter spectrograph since the
commissioning of the instrument, \startdate, and until the end of the last
period of the program 098.A-0055, \termdate~and thus constitutes eight years of GRB
afterglow observations with X-shooter. An overview of all the spectra and their
observational setups is given in Table \ref{tab:sample_overview}. For each
burst, each individual observation is provided in a separate reduction, and in
cases where observations have been repeated for an increased signal-to-noise or
to follow the temporal evolution, a combined spectrum is also provided. No
attempt has been made to join the spectroscopic arms, so for each observation,
three spectra are provided in separate files.

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=\columnwidth]{figures/timing.pdf}} \caption{Afterglow
	magnitude at the start of observation and redshift completeness as a function
	of follow-up delay for all the afterglows that have been followed up. The
	points have been colored based on the redshift of the corresponding burst. Red
	symbols indicate GRBs without a measured redshift and arrows indicate bursts
	for which the afterglow was not detected in the acquisition image. In red is
	shown the redshift completeness as a function of follow-up delay.}
\label{fig:timing}
\end{figure}

@@

\begin{figure*}
	\centerline{\includegraphics[width=\linewidth]{figures/completeness_BAT.pdf}}
\caption{Comparison between the burst properties of all bursts observed with
	\textit{Swift} (blue), the statistical sample that fullfill the criteria
	specified in Sect. \ref{samplecrit} (green), and the subset that has been
	observed as part of the statistical sample (red). The left plot shows the
	fluence in the 15-150 keV band as observed by BAT. The middle panel shows the 0.3 - 10
	keV flux, 11 hours after the bursts as measured by XRT. In the right most
	panel, we show $N_{\mathrm{HI,X}}$ based on the XRT spectrum
	\citep{Evans2009}.} \label{fig:swift_complete}
\end{figure*}

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=\columnwidth]{figures/XSGRB_p_values.pdf}}
\caption{Relational graph showing the respective $p$-values. They all represent
	the degree that the different samples are drawn from the same underlying
	distribution. The arrows represent the comparison direction with each of the
	$p$-values they are listed next to. Only in the H{\sc I} column density distribution is
	there mild evidence against the null hypothesis, but the discrepancy is mainly
	driven by a relatively larger fractional contribution from low-column hosts in
	the statistical sample.} \label{fig:p_values}
\end{figure}

@@

\begin{figure*}[!ht]
	\centering \includegraphics[width=\linewidth]{figures/XSGRB_redshift.pdf}
\caption{Redshift distribution as a function of intrinsic BAT $\gamma$-ray
	energy, $E_{\mathrm{BAT}}$. Bursts that are a part of the statistical sample
	are marked by blue stars whereas black dots show all GRBs observed with
	X-shooter. All \swift~GRBs with measured redshifts are shown in gray. For
	comparison, we overplot with red stars the GRB \citep{Tanvir2009b,
		Salvaterra2009a}, quasar \citep{Banados2017}, and galaxies \citep{Zitrin2015,
		Oesch2016} with the highest spectroscopically confirmed redshifts, the latter
	three shown at arbitrary $E_{\mathrm{BAT}}$. The blue solid line represents the
	so-called 1 s BAT sensitivity limit described in the text. The estimated epoch
	of reionization is shown by the black dot-dashed line, with the uncertainty
	shown as the black dashed lines, from the most recent measurement by
	\citet{Planck2015}. On the top of the plot is shown the marginalization of the
	redshift. Again the blue histogram represents the bursts than enter our sample
	and the white histogram the full GRB population. } \label{fig:z}
\end{figure*}

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=\columnwidth]{figures/betaOX.pdf}}
\caption{$\beta_{OX}$ against X-ray derived hydrogen column density. In red is
	shown the sample presented in \citet{Fynbo2009} where the lines indicate the
	kernel density estimate of the distribution. For the kernel density estimate,
	the limits have been replaced with values. We note that contrary to what is
	presented in \citet{Fynbo2009}, we exclude all bursts for which we do not have
	a redshift. Darker colors represent a higher density of points. In blue are the
	points for the bursts presented here along with the marginal histograms. Limits
	on $\beta_{OX}$ are shown by downwards facing arrows. The corresponding
	marginal distributions are shown along the edges of the plot. A Kendall's
	$\tau$ test, indicates a statistically significant, low degree of negative
	correlation ($\Gamma = -0.21$ at a $p$-value = 0.01)} \label{fig:betaOX}
\end{figure}

@@

\begin{figure}[!t]
	\centering \includegraphics[width=\columnwidth]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in Tanvir et al. (submitted), along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}.} \label{fig:NH_dist}
\end{figure}

@@

\begin{acknowledgements}
Much of the work done here would not have been done without Neil Gehrels, who has now passed away. We owe him a large debt for his invaluable work with the \swift~satellite. We also want to remember Javier Gorosabel, who was taken from us too early. 
%
JPUF, BMJ and DX acknowledge support from the ERC-StG grant EGGS-278202.
The Dark Cosmology Centre was funded by the Danish National Research
Foundation. 
%
This work was supported by a VILLUM FONDEN Investigator grant to JH
(project number 16599). 
%
TK acknowledges support by the European Commission
under the Marie Curie Intra-European Fellowship Programme in FP7.  
%
LK and JJ acknowledges
support from NOVA and NWO-FAPESP grant for advanced instrumentation in
astronomy. 
%
KEH and PJ acknowledge support by a Project Grant (162948--051) from
The Icelandic Research Fund. 
%
AG acknowledges the financial support from the
Slovenian Research Agency (research core funding No. P1-0031 and project grant
No. J1-8136). 
%
CT acknowledges support from a Spanish National Research Grant of Excellence
under project AYA 2014-58381-P and funding associated to a Ramón y Cajál
fellowship under grant number RyC-2012-09984.
%
AdUP acknowledges support from a Ramón y Cajal fellowship, a BBVA Foundation
Grant for Researchers and Cultural Creators, and the Spanish Ministry of Economy
and Competitiveness through project AYA2014-58381-P.
%
ZC acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2014-21669.
%
DAK acknowledges support from the Spanish research project AYA 2014-58381-P and
support from Juan de la Cierva Incorporaci\'on fellowships IJCI-2015- 26153.
%
RSR acknowledges AdUP's BBVA Foundation Grant for Researchers and Cultural
Creators and support from ASI (Italian Space Agency) through the Contract n. 2015-046-R.0 and from European Union Horizon 2020 Programme under the AHEAD project (grant agreement n. 654215).
%
GL is supported by a research grant (19054) from VILLUM FONDEN.
%
SDV acknowledges the support of the French National Research Agency (ANR) under contract ANR-16-CE31-0003 BEaPro
%
DM acknowledges support from the Instrument Center for Danish Astrophysics (IDA).
%
This work made use of data supplied by the UK {\it Swift} Science Data Centre at
the University of Leicester. Finally, it is with pleasure that we acknowledge
expert support from the ESO staff at the Paranal and La Silla observatories in
obtaining these target of opportunity data.
%
This research made use of Astropy, a community-developed core Python package for Astronomy \citep{TheAstropyCollaboration2013}. The analysis and plotting was achieved using the Python-based packages Matplotlib \citep{Hunter2007}, Numpy, and Scipy \citep{scipy, VanderWalt2011}, along with other community-developed packages.
%

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=\linewidth]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

These data have not been published elsewhere. Observations for GRB~161007A
started 323 hr after the burst trigger and were centered on the potential host
galaxy (see GCN 20014 \citep{GCN20014} and GCN 20020 \citep{GCN20020}). $4
\times 600$ s of observations were obtained revealing a faint continuum rising
abruptly above the noise at $\sim 685$ nm and extending up to 2100 nm. A very
low significance continuum is detected at shorter wavelengths, down to $\sim
600$ nm. No significant emission features could be identified. Two possibilities
can explain the observed continuum break, but neither are fully satisfactory.
The first option is the Lyman alpha cutoff at $z \sim 4.6$; at this redshift, the
usually detected strong nebular lines would be shifted out of the covered
wavelength range. The host would be however exceptionally bright, with an
absolute magnitude at $\approx 150$~nm{} $M = -23.8$~AB, that is about 2.8~mag
brighter than $M^*$ at that redshift (e.g. \citealt{Bouwens2015}). The inferred
X-ray hydrogen-equivalent column density would also be a record-high $4 \times
10^{23}$~cm$^{-2}$ (see e.g. \citealt{Campana2012}). Alternatively, the
continuum discontinuity could be the 400~nm{} break at $z = 0.71$; such
feature is however prominent only in early-type galaxies, which are unlikely
hosts of long-duration GRBs (but see \citealt{Rossi2014}). Given the lack of
optical variability, it cannot be excluded that the object within the XRT error
circle is a chance association, as the probability of a random alignment is
small but non negligible (of the order of 1\%). In conclusion, we cannot provide
a secure redshift for GRB 161007A.
